---
title: "Cardiovascular Risk Assessment RMD"
author: "Mark Klein"
date: "11/10/2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DATA CLEANING
## Read in packages
```{r}
library(ggplot2)
library(ROCR)
library(boot)
library(MASS)
library(ipred)
library(klaR)
library(ICS)
library(glmnet)
library(tree)
```

## Load and subset the data
```{r}
Data<-read.csv("health_failure_clinical_records_dataset.csv")
Data<-Data[,c(1,2,3,4,5,6,7,8,9,10,11,13)]
```

## Convert variables to factors and label
```{r}
Data$anaemia<-factor(Data$anaemia)
levels(Data$anaemia)<-c("Non-Anemic","Anemic")
Data$diabetes<-factor(Data$diabetes)
levels(Data$diabetes)<-c("Non-Diabetic", "Diabetic")
Data$high_blood_pressure<-factor(Data$high_blood_pressure)
levels(Data$high_blood_pressure)<-c("Non-Hypertensive", "Hypertensive")
Data$sex<-factor(Data$sex)
levels(Data$sex)<-c("Female", "Male")
Data$smoking<-factor(Data$smoking)
levels(Data$smoking)<-c("Non-Smoker", "Smoker")
Data$DEATH_EVENT<-factor(Data$DEATH_EVENT)
levels(Data$DEATH_EVENT)<-c(" No Death", "Death")
```


## Distribution of Response Variable
```{r}
ggplot(Data, aes(x=creatinine_phosphokinase)) + 
geom_histogram(aes(y=..density..), bins=10)+
labs(y="Frequency", x="Creatinine Phosphokinase (mcg/L)",
title="Distribution of Creatinine Phosphokinase")
```

## Change Response Variable in Data
```{r}
Data$creatinine_phosphokinase<-log(Data$creatinine_phosphokinase)
```

## Transformed Response Variable using Log Transformation
```{r}
ggplot(Data, aes(x=creatinine_phosphokinase)) + 
geom_histogram(aes(y=..density..), bins=10)+
labs(y="Frequency", x="Creatinine Phosphokinase (mcg/L)",
title="Distribution of Log(Creatinine Phosphokinase)")
```


## Split data training and test sets
```{r}
set.seed(0925)
sample.data<-sample.int(nrow(Data), floor(.50*nrow(Data)), replace = F)
train<-Data[sample.data, ]
test<-Data[-sample.data, ]
```


# EXPLORATORY DATA ANALYSIS

## Correlation Matrix of Selected Variables
```{r}
cor(train[,c(1,3,5,7,8,9)])
```

## Scatterplot Matrix of Selected Variables
```{r}
pairs(train[,c(1,3,5,7,8,9)], lower.panel=NULL)
```

## Scatterplot: Creatinine Phosphokinase vs. Age
```{r}
ggplot2::ggplot(train, aes(y=creatinine_phosphokinase, x=age))+
geom_point()+
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              geom = "smooth")+
labs(y="Creatine Phosphokinase (mcg/L)", x="Age (years)",
title="Scatterplot: Creatinine Phosphokinase vs. Age")
```

## Scatterplot: Creatinine Phosphokinase vs. Ejection Fraction
```{r}
ggplot2::ggplot(train, aes(y=creatinine_phosphokinase, x=ejection_fraction))+
geom_point()+
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              geom = "smooth")+
labs(y="Creatinine Phosphokinase (mcg/L)", x="Ejection Fraction (Percentage)",
title="Scatterplot: Creatinine Phosphokinase vs. Ejection Fraction")
```


## Scatterplot: Creatinine Phosphokinase vs. Platelets
```{r}
ggplot2::ggplot(train, aes(y=creatinine_phosphokinase, x=platelets))+
geom_point()+
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              geom = "smooth")+
labs(y="Creatinine Phosphokinase (mcg/L)", x="Platelets (kiloplatelets/mL)",
title="Creatinine Phosphokinase vs. Platelets")
```


## Scatterplot: Creatinine Phosphokinase vs. Serum Creatinine
```{r}
ggplot2::ggplot(train, aes(y=creatinine_phosphokinase, x=serum_creatinine))+
geom_point()+
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              geom = "smooth")+
labs(y="Creatinine Phosphokinase (mcg/L)", x="Serum Creatinine (mg/dL)",
title="Scatterplot: Creatinine Phosphokinase vs. Serum Creatinine")
```


## Scatterplot: Creatinine Phosphokinase vs. Serum Sodium
```{r}
ggplot2::ggplot(train, aes(y=creatinine_phosphokinase, x=serum_sodium))+
geom_point()+
  stat_smooth(method = "lm", 
              formula = y ~ x, 
              geom = "smooth")+
labs(y="Creatinine Phosphokinase (mcg/L)", x="Serum Sodium (mEq/L)",
title="Scatterplot: Creatinine Phosphokinase vs. Serum Sodium")
```


## Boxplot: Creatinine Phosphokinase vs. High Blood Pressure
```{r}
ggplot2::ggplot(train, aes(x=high_blood_pressure, y=creatinine_phosphokinase))+
geom_boxplot()+
labs(x="High Blood Pressure", y="Creatinine Phosphokinase (mcg/L)",
title="Boxplot of Creatinine Phosphokinase by High Blood Pressure")
```


## Boxplot: Creatinine Phosphokinase vs. Diabetes
```{r}
ggplot2::ggplot(train, aes(x=diabetes, y=creatinine_phosphokinase))+
geom_boxplot()+
labs(x="Diabetes", y="Creatinine Phosphokinase (mcg/L)",
title="Boxplot of Creatinine Phosphokinase by Diabetes")
```


## Scatterplot: Creatinine Phosphokinase vs. Anemia
```{r}
ggplot2::ggplot(train, aes(x=anaemia, y=creatinine_phosphokinase))+
geom_boxplot()+
labs(x="Anemia", y="Creatinine Phosphokinase (mcg/L)",
title="Boxplot of Creatinine Phosphokinase by Anemia")
```


## Scatterplot: Creatinine Phosphokinase vs. Sex
```{r}
ggplot2::ggplot(train, aes(x=sex, y=creatinine_phosphokinase))+
geom_boxplot()+
labs(x="Sex", y="Creatinine Phosphokinase (mcg/L)",
title="Boxplot of Creatinine Phosphokinase by Sex")
```

## Scatterplot: Creatinine Phosphokinase vs. Death Event
```{r}
ggplot2::ggplot(train, aes(x=DEATH_EVENT, y=creatinine_phosphokinase))+
geom_boxplot()+
labs(x="Death Event", y="Creatinine Phosphokinase (mcg/L)",
title="Boxplot of Creatinine Phosphokinase by Death Event")
```

# SHRINKAGE METHODS

## Prep Data for Shrinkage
```{r}
x<-model.matrix(creatinine_phosphokinase~.,data=Data)[,-c(1,12)]
head(x)
y<-Data$creatinine_phosphokinase
```

# RIDGE REGRESSION

## Setting the Threshhold
```{r}
ridge.r<-glmnet::glmnet(x,y,alpha=0, lambda=0)

## compare with OLS
result<-lm(creatinine_phosphokinase~age+anaemia+diabetes+ejection_fraction+high_blood_pressure+platelets+serum_creatinine+serum_sodium+sex+smoking,data=Data)
cbind(coefficients(result), coefficients(ridge.r))
```

```{r}
ridge.r<-glmnet::glmnet(x,y,alpha=0, lambda=0, thresh = 1e-23)
cbind(coefficients(result), coefficients(ridge.r))
```

## Split training and test
```{r}
set.seed(0925)
sample.data<-sample.int(nrow(Data), floor(.50*nrow(Data)), replace = F)
x.train<-x[sample.data,]
x.test<-x[-sample.data,]
y.train<-y[sample.data]
y.test<-y[-sample.data]
```

## Tuning Parameter
```{r}
set.seed(0925)
cv.out<-glmnet::cv.glmnet(x.train,y.train,alpha=0, thresh = 1e-23)
bestlam<-cv.out$lambda.min
bestlam
```

## Test MSE
```{r}
##fit ridge regression using training data and bestlam
ridge.mod<-glmnet::glmnet(x.train,y.train,alpha=0,lambda=bestlam,
thresh = 1e-23)

##Test MSE with best lambda
ridge.pred<-predict(ridge.mod,newx=x.test)
mean((ridge.pred-y.test)^2)
```

## Test MSE vs. Log(lambda)
```{r}
plot(cv.out)
```

# LASSO REGRESSION

```{r}
lasso.r<-glmnet::glmnet(x,y,alpha=1, lambda=0, thresh = 1e-23)
cbind(coefficients(result), coefficients(lasso.r))
```
## Tuning Parameter
```{r}
set.seed(0925)
cv.out.lasso<-glmnet::cv.glmnet(x.train,y.train,alpha=1, thresh = 1e-23)
bestlam.lasso<-cv.out.lasso$lambda.min
bestlam.lasso
```

## Test MSE
```{r}
lasso.mod<-glmnet::glmnet(x.train,y.train,alpha=1,lambda=bestlam.lasso, thresh = 1e-23)
lasso.pred<-predict(lasso.mod,newx=x.test)
mean((lasso.pred-y.test)^2)
```

## Test MSE vs. Log(lambda)
```{r}
plot(cv.out.lasso)
```

# ORDINARY LEAST SQUARES
```{r}
##fit OLS by setting lambda=0
ridge.mod.0<-glmnet::glmnet(x.train,y.train,alpha=0,lambda=0, thresh = 1e-23)

##test MSE with lambda=0
ridge.pred.0<-predict(ridge.mod.0,newx=x.test)
mean((ridge.pred.0-y.test)^2)
```


## REGRESSION TREE ##

# RECURSIVE BINARY SPLITTING

```{r}
tree.reg.train<-tree::tree(creatinine_phosphokinase~age+anaemia+diabetes+
                             ejection_fraction+high_blood_pressure+platelets
                           +serum_creatinine+serum_sodium+sex+smoking, data=train)
summary(tree.reg.train)
``` 

## Plot the tree
```{r}
plot(tree.reg.train)
text(tree.reg.train, cex=0.6, pretty=0)
```

## Test MSE
```{r}
pred.probs<-predict(tree.reg.train, newdata=test)
mean((pred.probs-y.test)^2)
```

# PRUNED TREE
```{r}
## use k=10 CV to prune
set.seed(2)
cv.class<-tree::cv.tree(tree.reg.train, K=10) 
## find size of tree that has the smallest deviance
trees.num.reg<-cv.class$size[which.min(cv.class$dev)]
trees.num.reg
```

## Plot pruned tree
```{r}
prune.reg<-tree::prune.tree(tree.reg.train, best=trees.num.reg)
plot(prune.reg)
text(prune.reg, cex=0.75, pretty=0)
```

## Summary of pruned tree
```{r}
summary(prune.reg)
``` 

## Test MSE 
```{r}
##prediction based on pruned tree for test data
tree.pred.reg<-predict(prune.reg, newdata=test)
# test MSE
mean((tree.pred.reg-y.test)^2)
```

# RANDOM FORESTS

## Fit regression tree with random forests
```{r}
set.seed(2)
rf.class<-randomForest::randomForest(creatinine_phosphokinase~age+anaemia+diabetes+ejection_fraction+high_blood_pressure+platelets+serum_creatinine+serum_sodium+sex+smoking, data=train, mtry=3,importance=TRUE)
summary(rf.class)
```


## Find test MSE
```{r}
# prediction based on random forest tree for test data
pred.rf<-predict(rf.class, newdata=test)
mean((pred.rf-y.test)^2)
```

## Find important predictors
```{r}
randomForest::varImpPlot(rf.class)
``` 

# RECURSIVE BINARY SPLITTING

## Data Cleaning
```{r}
Data$DEATH_EVENT <- as.factor(Data$DEATH_EVENT)
Data$creatinine_phosphokinase<-exp(Data$creatinine_phosphokinase)

```

## Split training and test
```{r}

set.seed(0925)
sample.data<-sample.int(nrow(Data), floor(.50*nrow(Data)), replace = F)
train<-Data[sample.data,]
test<-Data[-sample.data,]
y.test<-test[,"DEATH_EVENT"]
```

## Create tree
```{r}
tree.class.train<-tree::tree(DEATH_EVENT~., data=train)
summary(tree.class.train)
```
## Plot of tree
```{r}
plot(tree.class.train)
text(tree.class.train, cex=0.6, pretty=0)

```
## Confusion Matrix
```{r}
tree.pred.test<-predict(tree.class.train, newdata=test, type="class")
table(y.test, tree.pred.test)
```

## Overall test error rate
```{r}
accuracy <- mean(tree.pred.test==y.test)
test_error_rate <- 1 - accuracy
test_error_rate
```

## False Positive Rate
```{r}
FPR = 36 / (36+70)
FPR
```

## False Negative Rate
```{r}
FNR = 16 / (16+28)
FNR
```

## Changing threshold to 0.2 for confusion matrix
```{r}
#confusion matrix
tree.class.train<-tree::tree(DEATH_EVENT~., data=train)
pred.probs<-predict(tree.class.train, newdata=test)

table(y.test, pred.probs[,2]>0.2)

#accuracy
accuracy <- mean(tree.pred.test==y.test)
test_error_rate <- 1 - accuracy
test_error_rate
```

# PRUNED TREE

## Find optimal size of tree
```{r}
set.seed(0925)
cv.class<-tree::cv.tree(tree.class.train, K=10, FUN=prune.misclass)
trees.num.class<-cv.class$size[which.min(cv.class$dev)]
trees.num.class
```

## Create tree
```{r}
library(tree)
prune.class<-tree::prune.misclass(tree.class.train, best=trees.num.class)
prune.class
summary(prune.class)
```

## Plot of tree
```{r}
plot(prune.class)
text(prune.class, cex=0.75, pretty=0)

```

## Confusion Matrix
```{r}
tree.pred.test<-predict(prune.class, newdata=test, type="class")
table(y.test, tree.pred.test)
```

## Overall test error rate
```{r}
accuracy <- mean(tree.pred.test==y.test)
test_error_rate <- 1 - accuracy
test_error_rate
```

## False Positive Rate
```{r}
FPR = 28 / (28+78)
FPR

```

## False Negative Rate
```{r}
FNR = 11 / (11+33)
FNR

```

## Changing threshold to 0.1 for confusion matrix
```{r}
#confusion matrix
pred.probs<-predict(prune.class, newdata=test)
table(y.test, pred.probs[,2]>0.10)

#accuracy
accuracy <- mean(tree.pred.test==y.test)
test_error_rate <- 1 - accuracy
test_error_rate
```


# RANDOM FORESTS IMPROVEMENTS

## Create RF
```{r}
set.seed(2)
rf.class<-randomForest::randomForest(DEATH_EVENT~., data=train, mtry=3,importance=TRUE)
```
## Variable Importance
```{r}
randomForest::varImpPlot(rf.class)
```

## Predictions and confusion matrix for RF
```{r}
##prediction on test data
pred.rf<-predict(rf.class, newdata=test)
##confusion matrix for test data
table(y.test, pred.rf)
```

## Overall test error rate
```{r}
accuracy <- mean(pred.rf==y.test)
test_error_rate <- 1 - accuracy
test_error_rate
```

## False Positive Rate
```{r}
FPR = 20 / (20+86)
FPR
```

## False Negative Rate
```{r}
FNR = 14 / (14+30)
FNR
```

## Changing threshold to 0.3 for confusion matrix
```{r}
#confusion matrix
pred.probs<-predict(rf.class, newdata=test, type="prob")

table(y.test, pred.probs[,2]>0.3)
```